Plant species traits and growth rates: meta-analysis shows correlations change with plant size as predicted

### 2015-08-31 MS.tex
ms edition

\author[1]{Anais Gibert}
\author[1]{Emma F. Gray}
\author[1]{Mark Westoby}
\author[1]{Ian J. Wright}
\author[1]{Daniel S. Falster}
\affil[1]{Dept of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia}

@@

\textit{corresponding author} Anais Gibert, anais.gibert@gmail.com, Dept of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia, tel: +61 2 9850 8194

\textit{Statement of authorship} MW, DF, IW, AG designed the study, IW and DF provided their bibliography database, AG and EG searched and collected the data, AG performed the meta-analyses, DF performed modeling work, AG wrote the first draft, MW improved substantially the manuscript, and all authors contributed to revisions.

@@

\section*{Abstract}\label{abstract}

@@

\item 
We show that traits-GR correlations change in a predictable way as a function of plant size. Our understanding of plant strategies should shift away from attributing slow vs fast growth to species throughout life, in favour of attributing growth trajectories.

@@

\section*{Introduction}\label{introduction}
The phrase "plant ecological strategy" refers to how species face the challenge they all share:  to acquire sufficient water, nutrients and light for individual growth and to live long enough to replace themselves with offspring. An idea that species strategy differences might involve fast vs slow growth goes back at least to Grime and Hunt \citeyearpar{Grime:1975gr}, who measured a wide spread of relative growth rates for seedlings during early exponential growth under favourable conditions. Seedling potential relative growth rate (RGR, the change in a mass/height or diamater per time interval relative to initial size) has continued to be used as a major strategy indicator within the CSR (Competitor - Stress tolerator - Ruderal) scheme (e.g. \citealt{grime1979plant, Grime:1997wm}). Meanwhile partitioning of seedling RGR into its components showed that specific leaf area (SLA = 1/leaf mass per area LMA) was typically the strongest source of variation between species \citep{Poorter:1989tx}. From the 1990s a new style of "trait ecology" arose whereby measurable traits including SLA were used directly as strategy axes \citep{Westoby:2002ft}. An approach via measurable traits made broad comparisons possible across continents and latitudes and thousands of species. Traits have been seen as constructive approaches to understanding plant growth strategies \citep{Grime:1977kc,Chapin:1980gz}, community composition \citep{Lavorel:2002ff,Shipley:2006ie}, global vegetation dynamics \citep{Scheiter:2013ed} and ecosystem processes \citep{Lavorel:2002ff}.

Currently two main spectra of variation are widely cited as underpinning differences between species in vegetative growth rates. The 'leaf economic spectrum' \citep{Wright:2004jb} and the 'wood economic spectrum' \citep{Chave:2009iy} reflect tissue construction costs (for specific leaf area and for wood volume respectively) trading off with tissue turnover rates or mortality risks. It has been widely expected that species with low tissue construction cost will typically have fast growth rates at least in favourable physical environments (e.g. \citealt{MullerLandau:2004dc,Wright:2004jb,Poorter:2008iu,Chave:2009iy,Larjavaara:2010bn,Iida:2012jb,Paine:2015df}). Yet a growing number of empirical studies indicate the correlation between traits and growth is not as consistent as that. Consider SLA as an illustration. Species with high SLA can deploy more canopy area for a given dry mass investment in leaves \citep{Poorter:1999wd, Reich:1992wm}. In addition, SLA is correlated with high photosynthetic rates and high nitrogen (N) or phosphorus concentrations in the leaf \citep{Wright:2004jb}. For seedling relative growth rates under favourable conditions, high SLA has repeatedly been found to be a strong predictor regardless of vegetation type or growth form \citep{Lambers:1992bj,Reich:1992wm,Grime:1997wm,Poorter:1999wd,Wright:1999ds}. Yet for adult plants, this correlation appears weaker or absent \citep{coomes_comparison_1998,Poorter:2008iu,Aiba:2009ft,Easdale:2009gv,Wright:2010tp}. The growing discrepancy between theoretical expectation and empirical results has led some to question whether traits are even useful for understanding growth \citep{Wright:2010tp, Paine:2015df}.

Meanwhile the possibility has been raised that trait-growth correlations may change systematically with size \citep{Enquist:2007ek,Falster:2011ii, Ruger:2012jv, Iida:2014ep, Iida:2014hq}. Recent theory suggests that the absence of a correlation between SLA and growth at larger sizes may be expected, when potential influences of traits on plant function are considered \citep{Falster:2011ii, falster:2013}. At larger sizes the low-cost deployment benefits of high SLA may be offset or even outweighed by the high costs of rapid leaf turnover (see Text Box 1\ref{sec:growth}).

Here, we review published literature reporting on relationships between plant growth and five prominent traits (specific leaf area = SLA, wood or stem tissue density = WD, maximum leaf photosynthetic capacity per unit area = Aarea,  maximum height= Hmax, and seed mass= SM) in tree or woody species in order to assess the following questions.

\begin{enumerate}
  \item How general is the tendency for correlations between traits and growth rate to vary with plant size?
  \item Do the observed relationships conform to predictions (outlined below) from a recently developed  mechanistic
model \citep{Falster:2011ii, falster:2013}?
\end{enumerate}

While most studies on traits have not been concerned with effects of plant stage or size, they have reported trait-growth correlations for identifiable size classes. We gathered correlation coefficients between trait and growth from studies that gave explicit information about the size or stage of the plants. Meta-analysis allowed us to generalize about trait influences on growth.

@@

Recent theoretical work \citep{Falster:2011ii} allows hypotheses to be formulated about relationships between traits and growth, based on a mathematical decomposition of plant growth rate. Specific predictions arise from considering the influence of traits on elements of these equations and how these effects may vary with size (see Text Box 1\ref{sec:growth}).

@@

The correlation between trait and growth is predicted to revert according to the growth measurement used for both SLA and seed mass (see Text Box 1\ref{sec:growth}). Tree growth in diameter, height or mass can be expressed in either absolute (AbsGR) or relative terms (RGR). Here, the predicted relationship between growth and either SLA or seed mass in seedlings differs if absolute growth rate in mass is used instead of $RGR_{mass}$ (i.e. becomes negative for SLA, positive for SM).

@@

To test how general is the tendency for correlations between traits and growth rate to vary with plant size, we focused on an initial list of 20 traits (Fig. \ref{Fig1}). These traits have been selected because they capture different aspects of plant species strategies, and are commonly addressed in the functional traits litterature. Specific leaf area (SLA), leaf area, leaf nitrogen and phosphorus concentration, net (mass) assimilation rates (NAR), and assimilation rate of CO$_2$ (Aarea) are traits involved in species leaf economic strategy \citep{Wright:2004jb,Wright:2010tp}. Seed mass (SM), and maximum plant height (Hmax) capture information about reproductive strategy \citep{Falster:2005bw,Moles:2006ft}. Wood density (WD) and sapwood area per leaf area (SA/LA), vessel size and density, and stem specific conductance are all involved in species wood economic strategy \citep{Chave:2009iy}. Here WD and SA/LA are also considered as indicators of stem construction cost \citep{Falster:2011ii}, whereas SLA and leaf thickness capture crucial information about leaf construction cost \citep{Wright:2004jb}. Finally, the leaf area ratio and leaf mass ratio are indicators of resource allocation to leaf \citep{Lambers:1992bj}.

@@

We tested stage effect on traits for which > 15 correlation coefficients across at least two stages were recorded. We used a mixed-effect model of meta-analysis, in which it is assumed that studies within a stage share a common mean effect but that there is also random variation among studies in a stage, in addition to sampling variation. This random effect allows us to account for the heterogeneity among studies (e.g.~experimental type, vegetation type, error, etc.). For each trait, we established if stage could explain variance in the coefficient correlation by comparing models with the stage variable to the null model using log-likelihood ratios \citep{Zuur:2009cfa}. If the log-likelihood test returned a significant p-value, the stage model fitted the data better than the null model. Estimates of the effect size for each stage were considered significant if the 95\% confidence intervals did not overlap with zero. In the case of multiple comparisons within a study, we used the method proposed by \citet{Borenstein:2009um}: we calculated the mean correlation coefficient by trait and by study, and re-calculated the effect size. Following recent recommendations for the adoption of a checklist of quality criteria for meta-analysis in ecology (\citealt{Koricheva:2014ku}), we conducted several analyses to assess the reliability of our meta-analysis. Table A2 in Appendix 1 reported how and where (i.e in the text, appendix and/or figures) we addressed the 14 methodological quality criteria listed by \citet{Koricheva:2014ku}.

>>>>

\author[1]{Anais Gibert*}
\author[1]{Emma F. Gray}
\author[1]{Mark Westoby}
\author[1]{Ian J. Wright}
\author[1]{Daniel S. Falster}
\affil[1]{Dept of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia}

@@

\textit{Corresponding author} Anais Gibert, anais.gibert@gmail.com, Dept of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia, tel: +61 2 9850 8194

@@

\section*{Abstract (195/350 words)}\label{abstract}

@@

\item 
Synthesis: We show that traits-GR correlations change in a predictable way as a function of plant size. Our understanding of plant strategies should shift away from attributing slow vs fast growth to species throughout life, in favour of attributing growth trajectories.

@@

\section*{Introduction}\label{introduction}
Plant "ecological strategy" refers to how species face the challenge they all share: to acquire sufficient water, nutrients and light for individual growth and to live long enough to replace themselves with offspring. An idea that species strategy differences might involve fast vs slow growth goes back at least to Grime and Hunt \citeyearpar{Grime:1975gr}, who measured a wide spread of relative growth rates for seedlings during early exponential growth under favourable conditions. Seedling potential relative growth rate (RGR) has continued to be used as a major strategy indicator within the CSR (Competitor - Stress tolerator - Ruderal) classification (e.g. \citealt{grime1979plant, Grime:1997wm}). Meanwhile partitioning of seedling RGR into its components showed that specific leaf area (SLA = 1/leaf mass per area LMA) was typically the strongest source of variation between species \citep{Poorter:1989tx}. From the 1990s a new style of "trait ecology" arose whereby measurable traits including SLA were used directly as strategy axes \citep{Westoby:2002ft}. This approach via measurable traits made broad comparisons possible across continents and latitudes and thousands of species. Since then, traits have been seen as constructive approaches to understanding plant growth strategies \citep{Grime:1977kc,Chapin:1980gz}, community composition \citep{Lavorel:2002ff,Shipley:2006ie}, global vegetation dynamics \citep{Scheiter:2013ed} and ecosystem processes \citep{Lavorel:2002ff}.

Currently, two main spectra of variation are widely cited as underpinning differences between species in vegetative growth rates. The 'leaf economic spectrum' \citep{Wright:2004jb} and the 'wood economic spectrum' \citep{Chave:2009iy} reflect tissue construction costs (for specific leaf area and for wood volume respectively) trading off with tissue turnover rates or mortality risks. It has been widely expected that species with low tissue construction cost will typically have fast growth rates at least in favourable physical environments (e.g. \citealt{MullerLandau:2004dc,Wright:2004jb,Poorter:2008iu,Chave:2009iy,Larjavaara:2010bn,Iida:2012jb,Paine:2015df}). Yet, a growing number of empirical studies indicate the correlation between traits and growth is not as consistent as that. Consider SLA as an illustration. Species with high SLA can deploy more canopy area for a given dry mass investment in leaves \citep{Poorter:1999wd, Reich:1992wm}. In addition, SLA is correlated with high photosynthetic rates and high nitrogen (N) or phosphorus concentrations in the leaf \citep{Wright:2004jb}. For seedling relative growth rates under favourable conditions, high SLA has repeatedly been found to be a strong predictor regardless of vegetation type or growth form \citep{Lambers:1992bj,Reich:1992wm,Grime:1997wm,Poorter:1999wd,Wright:1999ds}. Yet for adult plants, this correlation appears weaker or absent \citep{coomes_comparison_1998,Poorter:2008iu,Aiba:2009ft,Easdale:2009gv,Wright:2010tp}. The growing discrepancy between theoretical expectation and empirical results has led some to question whether traits are even useful for understanding growth \citep{Wright:2010tp, Paine:2015df}.

Meanwhile the possibility has been raised that trait-growth correlations may change systematically with size \citep{Enquist:2007ek,Falster:2011ii, Ruger:2012jv, Iida:2014ep, Iida:2014hq}. Recent theory suggests that the absence of a correlation between SLA and growth at larger sizes may be expected, when potential influences of traits on plant function are considered \citep{Falster:2011ii, falster:2013}. At larger sizes the low-cost deployment benefits of high SLA may be offset or even outweighted by the high costs of rapid leaf turnover \citep{Falster:2011ii, falster:2013}. 
Recently, some empirical studies examined the relationships between functional traits and demographic rates across a wide range of tree sizes in specific experimental site (see \citealt{Iida:2014ep, Iida:2014hq}) and suggested a variation with plant size. However, it is still an opening question whether the influence of traits on growth change with plant size in a predictable and consistent way across different environemtal context.

In this study, we review published literature reporting on relationships between plant growth and five prominent traits (specific leaf area = SLA, wood or stem tissue density = WD, maximum leaf photosynthetic capacity per unit area = Aarea,  maximum height= Hmax, and seed mass= SM) in tree or woody species in order to assess the following questions.

\begin{enumerate}
  \item How general is the tendency for correlations between traits and growth rate to vary with plant size?
  \item Do the observed relationships conform to predictions (outlined below) from a recently developed mechanistic model \citep{Falster:2011ii, falster:2013}?
  \item
\end{enumerate}

While most studies on traits have not been concerned with effects of plant stage or size, they have reported trait-growth correlations for identifiable size/stage classes. Thus, we gathered correlation coefficients between traits and growth from studies that gave explicit information about the size or stage of the plants. Meta-analytic approach offers several advantage, in particular it allows us to generalize about the influence of traits on growth across different ecological context.

@@

Recent theoretical work \citep{Falster:2011ii} allows hypotheses to be formulated about relationships between traits and growth, based on a mathematical decomposition of plant growth rate. Specific predictions arise from considering the influence of traits on elements of these equations and how these effects may vary with size (see Appendix\ref{sec:growth} for mathematical explanations).

@@

The correlation between trait and growth is predicted to revert according to the growth measurement used for both SLA and seed mass (see Appendix 1\ref{sec:growth}). Tree growth in diameter, height or mass can be expressed in either absolute (AbsGR) or relative terms (RGR). Here, the predicted relationship between growth and either SLA or seed mass in seedlings differs if absolute growth rate in mass is used instead of $RGR_{mass}$ (i.e. becomes negative for SLA, positive for SM).

@@

We focused on an initial list of 20 traits (Fig. \ref{Fig1}). These traits have been selected because they capture different aspects of plant species strategies, and are commonly addressed in the functional traits litterature. Specific leaf area (SLA), leaf area, leaf nitrogen and phosphorus concentration, net (mass) assimilation rates (NAR), and assimilation rate of CO$_2$ (Aarea) are traits involved in species leaf economic strategy \citep{Wright:2004jb,Wright:2010tp}. Seed mass (SM), and maximum plant height (Hmax) capture information about reproductive strategy \citep{Falster:2005bw,Moles:2006ft}. Wood density (WD) and sapwood area per leaf area (SA/LA), vessel size and density, and stem specific conductance are all involved in species wood economic strategy \citep{Chave:2009iy}. Here WD and SA/LA are also considered as indicators of stem construction cost \citep{Falster:2011ii}, whereas SLA and leaf thickness capture crucial information about leaf construction cost \citep{Wright:2004jb}. Finally, the leaf area ratio and leaf mass ratio are indicators of resource allocation to leaf \citep{Lambers:1992bj}.

@@

We tested stage effect on traits for which > 15 correlation coefficients across at least two stages were recorded. We used a mixed-effect model of meta-analysis, in which it is assumed that studies within a stage share a common mean effect but that there is also random variation among studies in a stage, in addition to sampling variation. This random effect allows us to account for the heterogeneity among studies (e.g.~experimental type, vegetation type, error, etc.). For each trait, we established if stage could explain variance in the coefficient correlation by comparing models with the stage variable to the null model using log-likelihood ratios \citep{Zuur:2009cfa}. If the log-likelihood test returned a significant p-value, the stage model fitted the data better than the null model. Estimates of the effect size for each stage were considered significant if the 95\% confidence intervals did not overlap with zero. In the case of multiple comparisons within a study, we used the method proposed by \citet{Borenstein:2009um}: we calculated the mean correlation coefficient by trait and by study, and re-calculated the effect size. Following recent recommendations for the adoption of a checklist of quality criteria for meta-analysis in ecology (\citealt{Koricheva:2014ku}), we conducted several analyses to assess the reliability of our meta-analysis. Table 2 reported how and where (i.e in the text, appendix and/or figures) we addressed the 14 methodological quality criteria listed by \citet{Koricheva:2014ku}.
